Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-02 paper/paper.tex
Edits up to partway through 3.1

Our reduction and typing relations are defined in terms of \emph{simultaneous
substitutions} $[[rho]]$, which are mappings from variables to terms. We
use $[[idtm]]$ as the identity substitution, and the extension operator
$[[rho .: x -> a]]$ updates the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.
% When traversing under binders (e.g. in the
% $[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
% bound variable to itself and that the bound variable does not appear freely in
% the application of the substitution to any other variable.
%
\scw{Maybe we should handwave more here, and note that we aren't covering all
  of the details of the treatment of variable binding in the text. If readers
  want to understand that part, they should look at the Coq code, which uses
  de Bruijn indices anyways. Our goal is to convey the understanding of the
  Coq proof. }
\yl{I agree. Should mention upfront that the presentation is not
  watertight and rigorous, though we have one in our mechanization}
The substitution operator $[[a {rho}]]$ replaces every free variable $[[x]]$ of
$[[a]]$ by $[[rho x]]$ simultaneously. A single substitution corresponds to
composing extension and identity:
$[[a { b / x }]] \coloneqq [[a { idtm .: x -> b }]]$.

When reasoning about logical relations, it is more convenient to formulate
definitions in terms of simultaneous substitutions than single substitutions,
such as semantic typing in Section~\ref{sec:logrelproof}.
\jc{I've rewritten this paragraph to be more succinct,
but I'm still not certain it's necessary at all.
We're not focussing on the merits of simultaneous substitution over
other representations of substitution and binding, after all.}

@@

Before we define our logical relation, we first formally specify the
consistency property that we want to prove.
\begin{theorem}[Logical Consistency]
  \label{theorem:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{theorem}
The property can be formulated in a simply typed language, where
$[[Void]]$ is similarly defined as a type that has no term. A related
property, referred to as the termination property (for closed terms),
is commonly used in introductory materials such as
\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
\citet{harpertait} to motivate the need for a logical relation.

A naive attempt to proving Theorem~\ref{theorem:consistency} by
induction on the derivation $[[empty |- a : Void]]$ would succeed at
almost all cases except for \rref{T-App}. In the application
case, we are given $[[empty |- b : Pi x : A . B]]$ and $[[empty |- a : A]]$, and
the equality that $[[B {a / x} = Void]]$. Our goal is to show that
$[[empty |- b a : Void]]$ is not possible. However, note that there is
nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
because neither $[[Pi x : A . B]]$ nor $[[A]]$ is equal to $[[Void]]$.
We have no way of deriving a contradiction from $[[empty |- b a :
Void]]$. The takeaway from this failed attempt is that, in order to
derive the consistency, we need to know something about types other
than $[[Void]]$. From a pragmatic point of view, proof by logical
relation can be seen as a sophisticated way of strengthening the
induction hypothesis. From the strengthened property, the fundamental
theorem, we will be able to derive consistency as a corollary.

The complexity of applying proof by logical relation to dependent types stems
from the fact that the logical relation is much harder to define. In
simply typed languages, the logical relation is defined as a recursive
function over the type $[[A]]$.
In dependent types, the type
$[[A]]$ can take the form $[[(\ x . x ) Bool]]$. To assign meaning to
this type, we need to first reduce it to $[[Bool]]$. However, we
cannot write a function that performs the reduction because we do not
know the termination of well-typed terms a priori. As a result, we
define the logical relation as an inductively defined relation,
reminiscent of how we specify the reduction graph of a partial
function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.

\subsection{Definition of the Logical Relation}
\begin{figure}[H]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
\[ \drule[width=5in]{I-Pi} \]
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\[
    [[I]] \in [[ { j | j < i  } ->  PowerSet STm ]] \qquad\qquad\qquad
    [[S]] \in [[PowerSet STm]]
\]
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, the function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} ensures the predicativity of the system.

We tie the knot and obtain an interpretation for all universe levels
below. The judgment $[[InterpR i A S]]$ reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$. 
\begin{definition}[Logical relation for all universe levels]
\label{fig:logrelrec}
Define $[[InterpR i A S]]$ recursively through the well-foundedness of the $<$
relation on natural numbers.
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I j]]
    := [[{A | exists S , InterpR j A S}]] \text{ for } [[j < i]]
\end{equation*}
\end{definition}
Definition~\ref{fig:logrelrec} explains how the $[[j
< i]]$ constraint in \rref{I-Set} makes our system predicative; the
interpretation of the $[[i]]_{th}$ universe is only dependent on
universes strictly lower than $[[i]]$, which have already been defined.
This restriction ensures that that the relation is well-defined: without 
it the definition of $[[InterpR i A S]]$ would
not be well-founded; $[[Interp I i A S]]$ would call $[[I]]$ on
universe levels greater than or equal to $[[i]]$, which are yet to be defined.
%% SCW: removing the constraing wouldn't result in any system at all
% Removing the ordering constraint would result in a
% system where one can encode Girard's
% paradox~\citep{girard-thesis}.

@@

In most informal presentations, instead of defining the logical
relation in two steps as we have shown above, the rules for $[[InterpR
i A S]]$ are given directly, with the implicit understanding that the
relation is an inductive definition nested inside a recursive
function over the universe level $[[i]]$. We choose
the more explicit definition not only because it is directly definable
in proof assistants that lack induction-recursion,
% where inductive definitions must appear
%at the top level,
but also because it makes clear the induction
principle we are allowed to use when reasoning about $[[InterpR i A
S]]$.

% The paragraph below might be useful but I don't know how to phrase
% it well

@@

We next take a closer look at the inductive relation $[[Interp I i A S]]$, defined 
in Figure~\ref{fig:logrel}.
\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type only contains terms that
reduce to $[[true]]$ or $[[false]]$. Note that the characterization
of $[[Bool]]$ (and other inhabited types) in our logical relation does not always correspond to
well-typed or even closed terms. For example, the term $[[if false (#
Void true #)
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation when we extend our logical
relation to prove the existence of $\beta$-normal forms (Section~\ref{sec:extension}).
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}

>>>>

Our reduction and typing relations are defined in terms of \emph{simultaneous
substitutions} $[[rho]]$, which are mappings from variables to terms. Simultaneous
substitutions are more convenient to reason about than single substitutions,
especially when dealing with semantic typing in Section~\ref{sec:logrelproof}.
We use $[[idtm]]$ as the identity substitution, and the extension operator
$[[rho .: x -> a]]$ updates the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than to $[[rho x]]$.
% When traversing under binders (e.g. in the
% $[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
% bound variable to itself and that the bound variable does not appear freely in
% the application of the substitution to any other variable.
%
\scw{Maybe we should handwave more here, and note that we aren't covering all
  of the details of the treatment of variable binding in the text. If readers
  want to understand that part, they should look at the Coq code, which uses
  de Bruijn indices anyways. Our goal is to convey the understanding of the
  Coq proof. }
\yl{I agree. Should mention upfront that the presentation is not
  watertight and rigorous, though we have one in our mechanization}
The substitution operation $[[a {rho}]]$ replaces every free variable $[[x]]$ of
$[[a]]$ by $[[rho x]]$ simultaneously. A single substitution corresponds to
composing extension and identity:
$[[a { b / x }]] \coloneqq [[a { idtm .: x -> b }]]$.

@@

Our ultimate goal is to prove the following consistency property for our type theory.

\begin{theorem}[Logical Consistency]
  \label{theorem:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{theorem}
The property can be formulated in a simply typed language, where
$[[Void]]$ is a type with no constructors. A related
property, the \emph{termination property} (for closed terms),
is commonly used in introductory materials such as
\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
\citet{harpertait} to motivate the need for a logical relation.

A na\"ive attempt at proving Theorem~\ref{theorem:consistency} by
induction on the derivation $[[empty |- a : Void]]$ would succeed in
almost all cases except for \rref{T-App}. In the application
case, we are given $[[empty |- b : Pi x : A . B]]$ and $[[empty |- a : A]]$, and
the equality that $[[B {a / x} = Void]]$. Our goal is to show that
$[[empty |- b a : Void]]$ is not possible. Unfortunately, there is
nothing we know about $[[b]]$ or $[[a]]$ from the induction hypothesis
because neither $[[Pi x : A . B]]$ nor $[[A]]$ are equal to $[[Void]]$,
so we have no way of deriving a contradiction from $[[empty |- b a :
Void]]$. The takeaway from this failed attempt is that, in order to
derive the consistency, we need to know something about types other
than $[[Void]]$. From a pragmatic point of view, proof by logical
relation can be seen as a sophisticated way of strengthening the
induction hypothesis. From the strengthened property, the fundamental
theorem, we will be able to derive consistency as a corollary.

\jc{Minor comment but ``proof by logical relation'' sounds grammatically strange to me,
but I'm not sure whether ``proof by logical relations'' or
``proof by a logical relation'' would be any more correct...}

The challenge behind applying proofs by logical relation to dependent types
stems from the difficulty in defining the logical relation itself. In simply
typed languages, the logical relation is a recursive function over the type
$[[A]]$. In dependently typed languages, the type $[[A]]$ can take the form
$[[(\ x . x ) Bool]]$, for example. To assign meaning to this type, we need to
first reduce it to $[[Bool]]$. However, we cannot write a function that
performs the reduction because we do not know the termination of well-typed
terms a priori. As a result, we define the logical relation as an inductively
defined relation, reminiscent of how we specify the reduction graph of a
partial function.
We later recover functionality of the relation in Lemma~\ref{lemma:logreldeter}.

\subsection{Definition of the Logical Relation}

\begin{figure}
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
\[ \drule[width=5in]{I-Pi} \]
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}

The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation in Figure~\ref{fig:logrel}.
% The logical relation takes the form $[[Interp I i A S]]$.
The metavariables $[[A]]$ and $[[i]]$ are terms and naturals, respectively.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures,
using $[[PowerSet STm]]$ to denote the powerset of the set of terms.
\[
    [[I]] \in [[ { j | j < i  } ->  PowerSet STm ]] \qquad\qquad\qquad
    [[S]] \in [[PowerSet STm]]
\]
The function $[[I]]$ is a family of sets of terms indexed by natural numbers
strictly less than $[[i]]$, which represents the current universe level.
In \rref{I-Set}, the function $[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} ensures predicativity of the system.

Predicativity allows us to tie the knot and obtain an interpretation for all
universe levels as the judgment $[[InterpR i A S]]$, which says that the type
$[[A]]$ is a type at universe level $[[i]]$, \emph{semantically} inhabited by
terms from the set $[[S]]$.

\begin{definition}[Logical relation for all universe levels]
\label{fig:logrelrec}
$[[InterpR i A S]]$ is defined recursively by well-foundedness of the $<$
relation on natural numbers.
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I j]]
    := [[{A | exists S , InterpR j A S}]] \text{ for } [[j < i]]
\end{equation*}
\end{definition}

Our system is predicative because the interpretation of the $[[i]]$th universe
is dependent only on universes strictly below $[[i]]$,
which have already been defined.
This restriction ensures that the relation is well defined; otherwise, the
definition of $[[InterpR i A S]]$ would not be well founded, and
$[[Interp I i A S]]$ could call $[[I]]$ on universe levels greater than
or equal to $[[i]]$, which are yet to be defined.
%% SCW: removing the constraing wouldn't result in any system at all
% Removing the ordering constraint would result in a
% system where one can encode Girard's
% paradox~\citep{girard-thesis}.

@@

In most informal presentations, instead of defining the logical
relation in two steps as we have shown above, the rules for $[[InterpR
i A S]]$ are given directly, with the implicit understanding that the
relation is an inductive definition nested inside a recursive
function over the universe level $[[i]]$. We choose
the more explicit definition not only because it is directly definable
in proof assistants that lack induction--recursion,
but also because it makes clear the induction
principle we are allowed to use when reasoning about $[[InterpR i A S]]$.

@@

We next take a closer look at the remaining rules for the inductive relation
$[[Interp I i A S]]$ in Figure~\ref{fig:logrel}.
\Rref{I-Void, I-Bool} capture terms that \emph{behave} like the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type only contains terms that
reduce to $[[true]]$ or $[[false]]$. Note that the characterization
of $[[Bool]]$ (and other inhabited types) in our logical relation does not always correspond to
well-typed or even closed terms. For example, the term $[[if false (#
Void true #)
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation when we extend our logical
relation to prove the existence of $\beta$-normal forms (Section~\ref{sec:extension}).
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}
